---
title: "Assignment 2 Answers"
subtitle: ""
author: "Raghav Sinha, Yutong Han"
date: today
date-format: long
abstract: ""
number-sections: true
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
editor: 
  markdown: 
    wrap: 72
---

# Motorcycle Deaths

## Question 1

$$Y_t \sim \text{Poisson}(\mu_t)$$
$$log(\mu_t) = \beta_0 + f(t) + \text{offset}(\log(\text{MonthDays}_t)) + g(\text{month}_t) + \epsilon_t$$

-   $Y_t$ represents the weekly number of deaths at time t. It is
    assumed to follow a Poisson distribution.
-   The Poisson distribution is suitable as number of deaths is a count
    variable.
-   The log link function is used to connect the mean to the linear
    predictor. It ensures that the mean $\mu_t$ remains positive.
-   $\beta_0$ represents the baseline level of deaths when all other
    terms are zero.
-   $f(t)$ is a smooth function of time, it captures the non-linear
    trend in deaths over time.
-   $\log(\text{MonthDays}_t)$ is an offset that accounts for the
    varying number of days in each month.
-   $g(\text{month}_t)$ is a function that captures the seasonal effect
    of the month on the number of deaths (eg. more accidents in the
    summer due to increased riding.)

## Question 2

```{r}
#| include: false
#| warning: false
#| message: false
theUrl = "http://pbrown.ca/teaching/appliedstats/data/motorcycle.rds"
theFile = basename(theUrl)
if (!file.exists(theFile)) download.file(theUrl, theFile)
x = readRDS(theFile)
```

```{r}
#| warning: false
#| message: false
#| fig-width: 14
#| fig-height: 8
library(mgcv)
library(Hmisc)

x$dateInt <- as.integer(x$date)
x$logMonthDays <- log(Hmisc::monthDays(x$date))
x$month <- factor(format(x$date, "%b"), 
                  levels = format(ISOdate(2000, 1:12, 1), "%b"))

# Fit the GAM
gam_model <- gam(killed ~ s(dateInt, bs = "cr", k = 50) 
                 + offset(logMonthDays) + month,
                 data = x,
                 family = poisson(link = "log"),
                 method = "REML")

# Summary of the model
summary(gam_model)
```

```{r}
#| warning: false
#| message: false
#| include: false
# Generate predictions for new data
newdata <- data.frame(date = seq(as.Date("1975/1/1"), 
                                 as.Date("2030/1/1"), by = "month"))
newdata$dateInt <- as.integer(newdata$date)
newdata$logMonthDays <- log(30)
newdata$month <- format(newdata$date, "%b")
```

## Question 3

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 14
#| fig-height: 8
#| fig-cap: "Trend for for motorcycle deaths over time with 95% CI. The fitted curve captures both the long-term decline in deaths and a strong seasonal pattern. The parametric coefficients show significant seasonal effects, with deaths peaking in the summer months (June to August) and reaching a minimum in winter. For example, compared to January, deaths increase by 107% in August (exp(1.071) ≈ 2.92), while December shows no significant difference from January."
# Generate predictions with standard errors
pred <- predict(gam_model, newdata, type = "link", se.fit = TRUE)

# Compute confidence intervals (95% CI)
alpha <- 0.05  # Significance level for 95% confidence
z_value <- qnorm(1 - alpha / 2)  # Critical value for normal distribution
lower_ci <- pred$fit - z_value * pred$se.fit  # Lower bound of CI
upper_ci <- pred$fit + z_value * pred$se.fit  # Upper bound of CI

# Transform predictions and confidence intervals back to the response scale
pred_fit <- exp(pred$fit)  # Fitted values on the response scale
lower_ci <- exp(lower_ci)  # Lower CI on the response scale
upper_ci <- exp(upper_ci)  # Upper CI on the response scale

# Plot the observed data
plot(x$date, x$killed, cex = 0.2, log = "y", xlab = "Date", 
     ylab = "Number of Deaths",
     yaxt = "n")  # Suppress default y-axis labels

# Add custom y-axis labels
y_ticks <- c(5, 10, 20, 50, 100)
axis(2, at = y_ticks, labels = y_ticks, las = 1)

# Add fitted values and confidence bands
lines(newdata$date, pred_fit, col = "red", lwd = 2)  # Fitted values
polygon(c(newdata$date, rev(newdata$date)), c(lower_ci, rev(upper_ci)),
        col = rgb(1, 0, 0, 0.3), border = NA)  # Confidence band

# Add a legend
legend("topright", legend = c("Observed", "Fitted", "95% CI"),
       col = c("black", "red", rgb(1, 0, 0, 0.3)), lty = c(NA, 1, NA),
       pch = c(1, NA, NA), lwd = c(NA, 2, NA), 
       fill = c(NA, NA, rgb(1, 0, 0, 0.3)))
```

# Heat

```{r}
#| warning: false
#| message: false
#| include: false
library("mgcv")
# Additional code for downloading temperature data
heatURL = "http://pbrown.ca/teaching/appliedstats/data/sableIsland.rds"
dir.create("cache", showWarnings = FALSE)
heatFile = file.path("cache", basename(heatURL))
if (!file.exists(heatFile)) download.file(heatURL, heatFile)
x = readRDS(heatFile)
```

```{r}
#| message: false
#| warning: false
#| include: false
library(Pmisc)
library(GET) 
names(x) = gsub("[.]+C[.]", "", names(x))
x$Date = as.Date(x$Date)
x$month = as.numeric(format(x$Date, "%m"))
x$summer = x$month %in% 5:10

# Data transformation
x$dateInt = as.integer(x$Date)
x$yearFac = factor(format(x$Date, "%Y"))
xSub = x[x$summer & !is.na(x$Max.Temp), ]

# Define model file paths
model1_file <- "models/res1_model.rds"
model2_file <- "models/res2_model.rds"
model3_file <- "models/res3_model.rds"

# Model 1: Generalized Additive Model with k=100
if (file.exists(model1_file)) {
  res1 <- readRDS(model1_file)
} else {
  res1 = gam(
    update.formula(
      Max.Temp ~ s(dateInt, k = 100) + s(yearFac, bs = "re"),
      Pmisc::seasonalFormula(
        period = 365.25,
        harmonics = 1:2,
        var = "dateInt"
      )
    ),
    data = xSub,
    method = "ML",
    optimizer = "efs"
  )
  saveRDS(res1, model1_file)
}

# Model 2: Generalized Additive Model with k=4
if (file.exists(model2_file)) {
  res2 <- readRDS(model2_file)
} else {
  res2 = gam(
    update.formula(
      Max.Temp ~ s(dateInt, k = 4) + s(yearFac, bs = "re"),
      Pmisc::seasonalFormula(
        period = 365.25,
        harmonics = 1:2,
        var = "dateInt"
      )
    ),
    data = xSub,
    method = "ML",
    optimizer = "efs"
  )
  saveRDS(res2, model2_file)
}

# Model 3: Generalized Additive Model with k=100
if (file.exists(model3_file)) {
  res3 <- readRDS(model3_file)
} else {
  res3 = gam(
    update.formula(
      Max.Temp ~ s(dateInt, k = 100),
      Pmisc::seasonalFormula(
        period = 365.25,
        harmonics = 1:2,
        var = "dateInt"
      )
    ),
    data = xSub,
    method = "ML",
    optimizer = "efs"
  )
  saveRDS(res3, model3_file)
}
```

## Question 1

### Model 1 {.unnumbered}

$$
\text{Max.Temp}_i = \beta_0 + f_1(t_i) + b_{\text{year}_i} + \text{SEAS}(t_i) + \varepsilon_i
$$

$$
\begin{aligned}
    Y_i &\sim \mathcal{N} (\mu_i, \sigma^2) \\
    \mu_i &= f_2(t_i) + b_{\text{year}_i} 
    + \gamma_1 \sin\left( \frac{\pi t_i}{182.625} \right) 
    + \gamma_2 \cos\left( \frac{\pi t_i}{182.625} \right) \\
    &\quad + \gamma_3 \sin\left( \frac{\pi t_i}{91.3125} \right)
    + \gamma_4 \cos\left( \frac{\pi t_i}{91.3125} \right)
    + \varepsilon_i
\end{aligned}
$$

This model includes a flexible smooth trend over time using `s(dateInt, k
= 100)`, a random effect for each year using `s(yearFac, bs = "re")`, and
seasonal components using sine and cosine terms with annual and
semi-annual periods. It captures both long-term trends and year-to-year
variability. This is the most flexible and complex model. 

- ($Y_i$): daily maximum summer temperature, assumed Gaussian. 
- ($f_1(t_i)$): smooth function of time ($k=100$), capturing long-term temperature
trend. 
- ($b_{\text{year}_i} \sim \mathcal{N} (0, \sigma^2)$): random effect for year-level deviations. 
- Sine and cosine terms: seasonal components with 1-year and 6-month cycles. 
- ($\varepsilon_i$): residual error, absorbed into the Gaussian variance.

### Model 2 {.unnumbered}
$$
\text{Max.Temp}_i = \beta_0 + f_2(t_i) + b_{\text{year}_i} + \text{SEAS}(t_i) + \varepsilon_i
$$

$$
\begin{aligned}
    \text{Max.Temp}_i &= s(t_i, k = 4) + b_{\text{year}_i} 
    + \gamma_1 \sin\left( \frac{\pi t_i}{182.625} \right) 
    + \gamma_2 \cos\left( \frac{\pi t_i}{182.625} \right) \\
    &\quad + \gamma_3 \sin\left( \frac{\pi t_i}{91.3125} \right) 
    + \gamma_4 \cos\left( \frac{\pi t_i}{91.3125} \right) 
    + \varepsilon_i
\end{aligned}
$$ 
Model 2 (res2): This model is similar to res1, but the trend
component uses a much lower basis dimension k = 4. This means the
temporal trend is much smoother and less responsive to short-term
changes. It still includes the year random effect and seasonality.
Overall, it's a simpler and more conservative model, better for
detecting broad trends. 

- Identical structure to Model 1 but uses a
smoother trend term ($f_2(t_i)$) with fewer basis functions ($k=4$).

### Model 3 {.unnumbered}

$$
\text{Max.Temp}_i = \beta_0 + f_3(t_i) + \text{SEAS}(t_i) + \varepsilon_i
$$

$$
\begin{aligned}
    \text{Max.Temp}_i &= s(t_i, k = 100) 
    + \gamma_1 \sin\left( \frac{\pi t_i}{182.625} \right) 
    + \gamma_2 \cos\left( \frac{\pi t_i}{182.625} \right) \\
    &\quad + \gamma_3 \sin\left( \frac{\pi t_i}{91.3125} \right) 
    + \gamma_4 \cos\left( \frac{\pi t_i}{91.3125} \right) 
    + \varepsilon_i
\end{aligned}
$$
Model 3 (res3): This model keeps the flexible smooth trend (k = 100)
and seasonal components, but removes the year-level random effect. It
assumes that inter-annual variability is negligible or explained by the
overall trend. This model is simpler than res1, but potentially misses
year-specific deviations. 

- Removes the year-level random effect; assumes time trend alone captures variation.

## Question 2

Dear Editor,

While Ms. Burningier cites Figure 3(c) to claim "no clear evidence" of
warming, this interpretation reflects a critical misunderstanding of
statistical methodology. Model 3 omits year-level random effects—a
standard technique for addressing temporal autocorrelation in climate
data—resulting in excessive noise that obscures underlying trends.

## Question 3

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 14
#| fig-height: 8
#| fig-cap: "Panel A shows the estimated year-level random effects s(yearFac), which account for annual variability in maximum summer temperatures after removing the long-term trend and seasonal components. Panel B displays the estimated smooth temporal trend s(dateInt) with 200 posterior simulations (in gray) and an 80% joint confidence envelope (in red), revealing a clear long-term warming pattern consistent with climate change evidence."
Syear = unique(xSub$yearFac)
predYear = do.call(cbind, predict(res1,
                                  newdata = data.frame(yearFac = Syear, dateInt = 0),
                                  type = "terms", terms = "s(yearFac)", se.fit = TRUE)) %*%
  Pmisc::ciMat()

# Estimate Year Effects
newdat = data.frame(Date = seq(as.Date("1900/1/1"), as.Date("2035/12/31"), by = "2 weeks"),
                    yearFac = Syear[1])
newdat$dateInt = as.integer(newdat$Date)
predTrend = do.call(cbind, predict(res1, newdat,
                                   type = "terms", terms = "s(dateInt)", se.fit = TRUE)) %*%
  Pmisc::ciMat()

newX = predict(res1, newdata = newdat, type = "lpmatrix")
simCoef = rmvn(200, coef(res1), vcov(res1))  # 200 posterior samples
isTrend = grep("s[(]dateInt", colnames(newX))
simTrend = tcrossprod(newX[, isTrend], simCoef[, isTrend])  # Simulated trends

# Compute Smooth Trend
envV = GET::central_region(GET::curve_set(simTrend), coverage = 0.8)

# Panel A: Year Effects
Syear_numeric = as.numeric(as.character(Syear))
matplot(Syear_numeric, predYear, pch = 16, cex = 1.2,
        col = "black", xlab = "Year", ylab = "Estimated Year Effect (°C)",
        main = "Panel A: Estimated Year Effects s(yearFac) with 95% CI")
segments(Syear_numeric, predYear[, 2], Syear_numeric, predYear[, 3],
         lwd = 1.5, col = "gray40")
abline(h = 0, lty = 2, col = "darkgray")  # reference line

# Panel B: Smooth Trend
matplot(newdat$Date, simTrend, type = 'l', lty = 1,
        col = "#00000010", ylim = quantile(simTrend, c(0.01, 0.99)),
        xlab = "Date", ylab = "Estimated Smooth Trend s(dateInt) (°C)",
        main = "Panel B: Long-Term Smooth Trend with 80% Joint Confidence Region")
matlines(newdat$Date, envV[, c("lo", "hi")], lty = 1, col = "red", lwd = 2)
abline(h = 0, lty = 2, col = "darkgray")

# Optional legend
legend("topleft",
       legend = c("Simulated Trends", "80% Joint Confidence Region"),
       col = c("#00000050", "red"),
       lty = 1,
       lwd = c(1, 2),
       bty = "n",
       cex = 0.8)
```

## Question 4

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 14
#| fig-height: 8
# Filter summer data from 1900 to 1995
trainData = xSub[xSub$Date <= as.Date("1995-12-31"), ]
testData = xSub[xSub$Date >= as.Date("1996-01-01") & xSub$Date <= as.Date("2025-12-31"), ]

# Fix yearFac in test data — this is crucial!
testData$yearFac = "1995"

# Fit model using only data up to 1995 (same structure as res1)
resOld = gam(update.formula(
               Max.Temp ~ s(dateInt, k = 100) + s(yearFac, bs = "re"),
               Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
             ),
             data = trainData,
             method = "ML", optimizer = "efs")
# Predict
preds = predict(resOld, newdata = testData, se.fit = TRUE)
fitCI = cbind(preds$fit - 2 * preds$se.fit, preds$fit, preds$fit + 2 * preds$se.fit)
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 14
#| fig-height: 8
#| fig-cap: "*Panel A* shows observed summer maximum temperatures (gray dots) and the predicted trend (blue line) from a GAM model trained only on data up to 1995. The model clearly captures the warming trend observed between 1996 and 2025, supporting its long-term forecasting ability. *Panel B* plots observed vs. predicted maximum summer temperatures. The strong alignment around the 1:1 line confirms the model’s predictive accuracy, even for future data it has never seen."

par(mfrow = c(1,2))  # Arrange plots side by side

# Panel A: Observed vs Forecasted Trend
plot(testData$Date, testData$Max.Temp, col = "gray70", pch = 16,
     xlab = "Date", ylab = "Max Temp (°C)",
     main = "Panel A: Observed vs Forecasted Trend (1996–2025)")
lines(testData$Date, preds$fit, col = "blue", lwd = 2)
matlines(testData$Date, fitCI[, c(1, 3)], col = "blue", lty = 2)

# Panel B: Observed vs. Predicted
plot(testData$Max.Temp, preds$fit,
     pch = 16, cex = 0.6, col = "darkblue",
     xlab = "Observed Max Temp (°C)",
     ylab = "Predicted Max Temp (°C)",
     main = "Panel B: Observed vs. Predicted Max Summer Temperatures (1996–2025)")
abline(0, 1, col = "red", lty = 2, lwd = 2)

par(mfrow = c(1,1))  # Reset plot layout
```

The claim that climate models “have consistently failed to predict the
future” is not supported by these results.\
A GAM model trained only on pre-1996 data successfully captures both the
upward trend and year-to-year variability in summer temperatures from
1996 to 2025.\
This demonstrates that well-specified statistical models can provide
meaningful and accurate climate forecasts over multi-decade horizons.
