---
title: "Assignment 2 Answers"
subtitle: ""
author: "Raghav Sinha, Yutong Han"
date: today
date-format: long
abstract: ""
number-sections: true
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
editor: 
  markdown: 
    wrap: 72
---
# Motorcycle Deaths

## Question 1
$$Y_t \sim \text{Poisson}(\mu_t)$$
$$log(\mu_t) = \beta_0 + f(t) + \text{offset}(\log(\text{MonthDays}_t)) + g(\text{month}_t) + \epsilon_t$$

- $Y_t$ represents the weekly number of deaths at time t. It is assumed to follow a Poisson distribution.
- The Poisson distribution is suitable as number of deaths is a count variable. 
- The log link function is used to connect the mean to the linear predictor. It ensures that the mean $\mu_t$ remains positive.
- $\beta_0$ represents the baseline level of deaths when all other terms are zero.
- $f(t)$ is a smooth function of time, it captures the non-linear trend in deaths over time.
- $\log(\text{MonthDays}_t)$ is an offset that accounts for the varying number of days in each month.
- $g(\text{month}_t)$ is a function that captures the seasonal effect of the month on the number of deaths (eg. more accidents in the summer due to increased riding.)

## Question 2

```{r}
#| include: false
#| warning: false
#| message: false
theUrl = "http://pbrown.ca/teaching/appliedstats/data/motorcycle.rds"
theFile = basename(theUrl)
if (!file.exists(theFile)) download.file(theUrl, theFile)
x = readRDS(theFile)
```

```{r}
#| warning: false
#| message: false
#| fig-width: 14
#| fig-height: 8
library(mgcv)
library(Hmisc)

x$dateInt <- as.integer(x$date)
x$logMonthDays <- log(Hmisc::monthDays(x$date))
x$month <- factor(format(x$date, "%b"), 
                  levels = format(ISOdate(2000, 1:12, 1), "%b"))

# Fit the GAM
gam_model <- gam(killed ~ s(dateInt, bs = "cr", k = 50) 
                 + offset(logMonthDays) + month,
                 data = x,
                 family = poisson(link = "log"),
                 method = "REML")

# Summary of the model
summary(gam_model)
```

```{r}
#| warning: false
#| message: false
#| include: false
# Generate predictions for new data
newdata <- data.frame(date = seq(as.Date("1975/1/1"), 
                                 as.Date("2030/1/1"), by = "month"))
newdata$dateInt <- as.integer(newdata$date)
newdata$logMonthDays <- log(30)
newdata$month <- format(newdata$date, "%b")
```

## Question 3

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 14
#| fig-height: 8
#| fig-cap: "Trend for for motorcycle deaths over time with 95% CI. The fitted curve captures both the long-term decline in deaths and a strong seasonal pattern. The parametric coefficients show significant seasonal effects, with deaths peaking in the summer months (June to August) and reaching a minimum in winter. For example, compared to January, deaths increase by 107% in August (exp(1.071) ≈ 2.92), while December shows no significant difference from January."
# Generate predictions with standard errors
pred <- predict(gam_model, newdata, type = "link", se.fit = TRUE)

# Compute confidence intervals (95% CI)
alpha <- 0.05  # Significance level for 95% confidence
z_value <- qnorm(1 - alpha / 2)  # Critical value for normal distribution
lower_ci <- pred$fit - z_value * pred$se.fit  # Lower bound of CI
upper_ci <- pred$fit + z_value * pred$se.fit  # Upper bound of CI

# Transform predictions and confidence intervals back to the response scale
pred_fit <- exp(pred$fit)  # Fitted values on the response scale
lower_ci <- exp(lower_ci)  # Lower CI on the response scale
upper_ci <- exp(upper_ci)  # Upper CI on the response scale

# Plot the observed data
plot(x$date, x$killed, cex = 0.2, log = "y", xlab = "Date", 
     ylab = "Number of Deaths",
     yaxt = "n")  # Suppress default y-axis labels

# Add custom y-axis labels
y_ticks <- c(5, 10, 20, 50, 100)
axis(2, at = y_ticks, labels = y_ticks, las = 1)

# Add fitted values and confidence bands
lines(newdata$date, pred_fit, col = "red", lwd = 2)  # Fitted values
polygon(c(newdata$date, rev(newdata$date)), c(lower_ci, rev(upper_ci)),
        col = rgb(1, 0, 0, 0.3), border = NA)  # Confidence band

# Add a legend
legend("topright", legend = c("Observed", "Fitted", "95% CI"),
       col = c("black", "red", rgb(1, 0, 0, 0.3)), lty = c(NA, 1, NA),
       pch = c(1, NA, NA), lwd = c(NA, 2, NA), 
       fill = c(NA, NA, rgb(1, 0, 0, 0.3)))
```